Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-02 doc/papers/cgo2014/paper.tex
CGO: partial review

Fortran/C/C++ have been at the hearth of high performance computing for many years.
Compiler infrastructures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
good abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of third-party libraries like GotoBlas~\cite{gotoblas2008}.
% pas de nt2?

There is no need to reinvent the wheel when compiling numerical Python kernels.
A more profitable approach is to map Python concept to existing ones, or
translate them into existing them. As a consequence, this section focuses on
the translation of Python constructs to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.

@@

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the Rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
typedef typename x::dtype T;
constexpr size_t N = x::ndim;
ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

The computation of read-write effects involve the annotation of Numpy functions
so that the Python compiler knows that most of them do not modify their
arguments, and the computation of argument effects for user functions. This can
be done by considering each function individually and looking for the two kind
of constructs that can modify a function argument: indexing and further
function calls.

To inter procedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions. Each node contains a list of integer representing the indices of the
formal argument that are modified intra-procedurally. Then we add an edge
between two nodes, labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
writes $i$, any outcoming edge $(i, j)$ can be safely removed, as we know for
sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
$(j, i)$ can be removed and $j$ is added to the source's list. Iteratively
applying this process removes all edges that can propagate a write, the only
remaining edges propagate a read and can be ultimately removed.

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that writes \texttt{n} through a
subscript. Outcoming edges (1) and (2) can be removed because they are
outcoming edges of \texttt{bar} starting by \texttt{n}. In coming edge (0) ends
by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
know that both \texttt{root} update \texttt{r} and \texttt{bar} update \texttt{n},
but \texttt{foo} doesn't update its argument.

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its argument or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially easy in Python thanks to the
ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate these optimisation. It is use with Pythran (commit: 488b7ba)
with optimization enabled/disabled. Finally, we compare result with Pypy (???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

@@

% For performance comparison between different Pythran version
To compare optimisation result, we run the numpy-benchmarks test suite multiple
time on the rosen.py benchmark. Results on Figure~\ref{rosenamd48.png} and \ref{rosenintel8.png} show difference
one speed which could be explain by AVX support on the intel8 architecture while amd48
support up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.

>>>>

Fortran/C/C++ have been at the hearth of high performance computing for many years.
Compiler infrastructures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit the need of hand-written assembly to a few hot 
spots.
Language extensions like OpenMP~\cite{openmp4} provide a basic support for 
parallel programming, while libraries like Boost.SIMD~\cite{esterie2012boost} 
offer good abstractions for explicit vectorization. Many highly tuned kernels 
are available in the form of third-party libraries like 
GotoBlas~\cite{gotoblas2008}.
% pas de nt2?

There is no need to reinvent the wheel when compiling numerical Python kernels.
A profitable approach is to map Python high-level to existing ones, or
translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
As a consequence, this section focuses on
the translation of Python constructs to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.

@@

\begin{lstlisting}[language=c++, 
  caption={C++11 translated version of Python version for the Rosenbrock 
           kernel.}, 
  label={lst:rosen-cxx}, 
  breaklines=true,
  basicstyle=\small]
typedef typename x::dtype T;
constexpr size_t N = x::ndim;
ndarray<T, N> t0 = 100L * 
  numpy::square{}(x[slice(1L, None)] - 
  numpy::square{}(x[slice(None,-1L)]));
ndarray<T, N> t1 = 
  numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

The computation of read-write effects involves the computation of argument 
effects for user functions, and the annotation of Numpy functions
so that the Python compiler knows that most of them do not modify their
arguments. This can
be done by considering each function individually and looking for the two kind
of constructs that can modify a function argument: indexing and further
function calls.

To inter procedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions. Each node contains a list of integer representing the indices of the
formal argument that are modified intra-procedurally. Then we add an edge
between two nodes, labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
writes $i$, any outgoing edge $(i, j)$ can be safely removed, as we know for
sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
$(j, i)$ can be removed and $j$ is added to the source's list. Iteratively
applying this process removes all edges that can propagate a write, the only
remaining edges propagate a read and can be ultimately removed.

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that writes \texttt{n} through a
subscript. Outgoing edges (1) and (2) can be removed because they are
outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
know that both \texttt{root} update \texttt{r} and \texttt{bar} update \texttt{n},
but \texttt{foo} doesn't update its argument.

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its arguments or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially easy in Python thanks to the
ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate these optimizations. It is use with Pythran (commit: 488b7ba)
with optimization enabled/disabled. Finally, we compare result with Pypy (???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test 
suite multiple time on the \textit{rosen} benchmark.
Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
The difference between the two architectures can be explain by AVX support on 
the Intel8 architecture while AMD48 support up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.
